The interactions of rational, pragmatic agents  \n lead to efficient language structure and use

### 2019-02-01 paper/paper.Rmd
1st round FD updates

Zipf articulated what is now considered a *functionalist* approach to language science -- analyzing language structure and use in terms of efficiency. Such an approach might reframe our opening question as follows: how does having property \textit{x} make using language $\ell$ more or less useful for communication? Among others, this efficiency-based framing has produced a rich set of theoretical and empirical targets exploring semantic typology [@RegierKempKay2015a], properties such as ambiguity [@Piantadosi2011a] and compositionality [@KirbyGriffithsSmith2014a], and the efficient use of reduction and redundancy in production [@LevyJaeger2007a; @GenzelCharniak2002a]. These projects typically employ information-theoretic principles to characterize efficiency in natural or simulated language corpora. We adopt a different approach in this work, deriving a measure of efficiency from first principles of \textit{efficient language use} and show how this can lead to communicatively efficient systems.\par

@@

Adopting reference games as a basic unit of analysis suggests that optimization may take place at the level of conversation. Importantly, Zipf's conception of speaker and listener effort may be connected to how language is used; in particular, whether interlocutors engage in pragmatic reasoning during conversation. Under a Gricean treatment of pragmatics, speakers and listeners follow a set of conversational maxims in which they cooperate to transfer information [@Grice1975a]. These maxims appear to emerge from efficiency concerns, however (Horn, 1984). We formalize this connection -- showing how system-level efficiencies can emerge from local interaction behavior of pragmatic agents. Intuitively, our claim is that to understand an "efficient" property of a system it is essential that we consider how that property is *used* efficiently.\par

@@

### Speakers and listeners
The Rational Speech Act framework (RSA) is a computational-level theory of pragmatic language use, which has produced good fit to human communication behavior across a range of language phenomena [@FrankGoodman2012a; @GoodmanFrank2016a]. RSA is a formalization of essential Gricean pragmatic principles -- agents reason about one another and their shared context [@Grice1975a]. We adopt RSA as our representational framework to model Gricean (rational and pragmatic) speaker-listeners in the reference game setting (see SI and @GoodmanFrank2016a for review).\par

@@

Importantly, we assume that events $e = <u, m, c>$ are sampled according the to following generative model -- some context occurs in the world with probability $P(C=c)$. Within this context, an object $m$ occurs with probability $p(m|c)$. The speaker attempts to refer to that object by sampling from her conditional distribution $S(u|m, c; \ell)$ (i.e. $e \sim p(c)p(m|c)S(u|m, c; \ell)$). From the ingredients described above, it is possible to derive the following objective between the speaker and listener joint distributions (see SI 2.2 for a complete derivation).

@@

From an information-theoretic perspective this objective is intuitive: $H_{cross}$ denotes the Cross-Entropy (CE), a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, we have an expectation over this term -- the expected difference between the distributions assumed by the speaker $P_{speaker}$ and listener $P_{listener}$ given a set of contexts $C$\footnote{Note that in the single context case $|C|=1$ this objective is simply the speaker-listener Cross-Entropy over utterance-meaning pairs $H_{cross}[P_{speaker}(u, m; \ell), P_{listener}(u, m; \ell)]$}. In other words, an "efficient" language $\ell$ minimizes the distance between what speakers and listeners think.

@@

The task of understanding language is marked by a frequent need to handle various forms of ambiguity: lexical, syntactic, among others [@WasowPerforsBeaver2005a]. The ubiquity of this property, however, has been argued to provide evidence that languages have not been optimized for communication\footnote{Chomsky's (2002) claim that "... language is not well designed, because you have such properties as ambiguity" is analogous to the Zipfian optimal Listener language $\ell_{listener}^*$.} [@Chomsky2002a].

@@

In the following experiments we explore two aspects of Piantadosi’s et al.'s claim. In Simulation 1, we examine the efficient language *structure* aspect of their claim, exploring when the optimal linguistic system $\ell^*$ is most likely to contain ambiguous lexical items. In Simulation 2, we explore an efficient language *use* aspect of the claim -- at what point in a conversation is it useful for a speaker to use ambiguous lexical material?\par

```{r plot-optimal-langs, fig.env = "figure*", fig.pos = "h", fig.width=6, fig.height=3, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Panel (A) Vertical axis shows the proportion of optimal languages containing ambiguity. Horizontal axis shows the number of context-sizes (1-4). Optimal language under CE objective (red). Speaker-optimal (blue). Listener-optimal (green). Error bars represent 95 percent confidence intervals. Panel (B), example CE-optimal language (ambiguous) under in a four-condition context. Panel (C), example CE-optimal language in a single-condition context."}
img <- png::readPNG("figs/fig1.png")
grid::grid.raster(img)
```

@@

We show that ambiguity is an efficient property under our CE objective in the reference game setting. We proceed by generating languages with different amounts of contextual support (varying the size of $|C|$). We search the space of languages, examining whether ones which minimize our objective contain ambiguity. If context leads to more efficient communication, then optimal languages under our objective should be more likely to be ambiguous as the amount of context increases.\par

@@

For comparison, we also examine properties of optimal languages under two additional objectives. Zipf [-@Zipf1949a] proposed that the speaker-optimal language $\ell_{speaker}^*$ would minimize speaker effort and the listener-optimal language $\ell_{listener}^*$ would minimize listener effort. We define these objectives using the first and second half of equation 1 (see SI section 2.2. for more detail).\par

@@

In Simulation 1 we explored the degree to which ambiguity is an efficient property of languages when communication is contextualized. Figure 1 plots the proportion of optimal languages under each objective as a function of number of contexts. The red line shows that as the number of contexts increases, so does the probability that an optimal language $\ell^*_{cross}$ contains ambiguity (at least one utterance maps to two meanings) under our CE objective. For comparison we also plot the proportion of speaker-optimal $\ell^*_{speaker}$ (blue line) and listener-optimal $\ell^*_{listener}$ (green line) languages that contain ambiguity. In line with Zipf's predictions, if languages are designed only to minimize speaker effort then optimal languages always contain ambiguity. If languages are designed to minimize listener effort then ambiguity is always avoided.\par

@@

We consider consider a single language $\ell$, which contains both ambiguous and unambiguous utterances. We assume ambiguous utterances are lower cost. Crucially, we do not assume that the listener knows the particular topic ($c_{current}$) of the conversation \textit{a priori}. Rather, that the listener has knowledge of the set of possible topics $C = \{c_1, \dots, c_k\}$, but does not know which one is currently being used by the speaker. Formally, this means the listener does not have access to the correct conditional distribution over meanings $P(M|C=c_{current})$ at the start of the discourse.\par

Over the course of a discourse $D$, the listener tries to infer both the current topic, $c_{current}$, as well as the particular meaning $m$ of a given utterance $u$. That is, we consider listener and speaker agents who can track the history of previous utterances $D$. Importantly, an agent can attempt to infer the current topic of conversation $c_{current}$ using the discourse history $D$.\par

@@

## Speaker agents
We vary the degree to which agents can use context for disambiguation. We consider three types of speaker models. Our \textit{Full pragmatics} agent, models a speaker who reasons about her listener and also has complete recall of the set of utterances in the discourse $D$. This speaker believes that the listener may not know the current topic $c_{current}$ at the start of the discourse, but can infer it over the discourse. We compare two baseline models. The first, a \textit{Partial pragmatics} baseline describes a speaker who reasons about a listener, but assumes they have no access to the discourse history. The second, a \textit{No pragmatics} speaker does not consider a listener at all, but produces utterances according to the underlying langauge semantics ($\ell$) and topic probabilities ($p(M|C=c_{current}$) (see SI section 3).\par

@@

Figure 2 (A) shows the empirical probability that a speaker uses an ambiguous utterance as a function of discourse position. The \textit{No pragmatic} baseline uses ambiguous utterances frequently and at a constant rate over the discourse and the \textit{Partial pragmatic} baseline avoids ambiguous utterances entirely. But, the \textit{Full pragmatic} model avoids ambiguous material only at the start of the discourse, employing it increasingly as the discourse proceeds. Panel (C) tracks our CE objective for each model over the discourse. Note that the objective decreases for all three models, primarily driven by the listener updating his belief about the actual topic ($P(C=c_{current}|D)$). However, the objective declines more quickly under the \textit{Full} and \textit{Partial} pragmatic speakers as listener agents are better able to infer the correct context. Additionally, the difference in CE between the \textit{Full}- and \textit{Partial pragmatic} models at the end of the discourse is driven by the reduction in speaker costs (see panel B). While speaker effort remains constant in both \textit{No pragmatic} and \textit{Partial pragmatic} baselines, effort declines in the \textit{Full pragmatic} model as she increasingly relies on ambiguous material later in the discourse.\par

@@

How do the competing pressures of speakers and listeners give rise to the distributional regularities found in natural language? Zipf [-@Zipf1949a] proposed that the asymmetry between speaker and listener costs gives rise to a range of properties at the level of the lexicon. We explored the interactions of rational pragmatic agents as a framework for understanding efficient language structure and use. We focused on an argument on the communicative function of ambiguity [@Piantadosi2011a], deriving a novel speaker-listener Cross-Entropy objective for measuring the efficiency of linguistic systems from first principles of efficient language use. In a series of simulations we showed that optimal languages are more likely to contain ambiguous material when context is informative and that rational pragmatic agents will use ambiguous material efficiently.\par

A limitation of the current work is an analysis of exactly how this objective compares to existing measures. For example, previous work has described competing speaker-listener pressures in terms of a trade-off of simplicity and informativeness [@KempRegier2012a] or expressivity and compressibility [@SmithTamarizKirby2013a]. Future work should assess the degree to which we can derive the same properties as previous studies using our current framework. In line with this idea, a primary focus of the current work is to provide a case-study on the productivity of assessing functionalist theories in repeated reference games between rational pragmatic agents. We hope this framework can serve as a domain general tool to assess the range of functionalist theories examining efficient language-structure and use.\par

>>>>

Zipf articulated what is now considered a *functionalist* approach to language science -- analyzing language structure and use in terms of efficiency. Such an approach might reframe our opening question as follows: how does having property \textit{x} make using language $\ell$ more or less useful for communication? Among others, this efficiency-based framing has produced a rich set of theoretical and empirical targets exploring semantic typology [@RegierKempKay2015a], properties such as ambiguity [@Piantadosi2011a] and compositionality [@KirbyGriffithsSmith2014a], and the efficient use of reduction and redundancy in production [@LevyJaeger2007a; @GenzelCharniak2002a]. These projects typically employ information-theoretic principles to characterize efficiency in natural or simulated language corpora.\par

We adopt a different approach in this work in two ways. First we derive a measure of efficiency from first principles of \textit{efficient language use} and show how this can lead to communicatively efficient systems. Second we demonstrate how assumptions about language-users impacts whether language properties are used efficiently. In this way, we bridge questions of design and use in a single framework.\par

@@

Adopting reference games as a basic unit of analysis suggests that optimization may take place at the level of conversation. Importantly, Zipf's conception of speaker and listener effort may be connected to how language is used; in particular, whether interlocutors engage in pragmatic reasoning during conversation. Under a Gricean treatment of pragmatics, speakers and listeners follow a set of conversational maxims in which they cooperate to transfer information [@Grice1975a]. These maxims appear to emerge from efficiency concerns, however (Horn, 1984). We formalize this connection -- showing how system-level efficiencies can emerge from local interaction behavior of pragmatic agents. Our claim is that to understand an "efficient" property of a system it is essential that we consider how that property is *used* efficiently.\par

@@

### Speakers and listeners
The Rational Speech Act framework (RSA) is a computational-level theory of pragmatic language use, which has produced good fit to human communication behavior across a range of language phenomena [@FrankGoodman2012a; @GoodmanFrank2016a]. RSA is a formalization of essential Gricean pragmatic principles -- agents reason about one another and their shared context [@Grice1975a]. We adopt RSA as our representational framework to model Gricean (rational and pragmatic) speaker-listeners in the reference game setting (see \href{https://github.com/benpeloquin7/zipf_principles/blob/master/paper/supplementary_materials.pdf}{SI} and @GoodmanFrank2016a for a review).\par

@@

Importantly, we assume that events $e = <u, m, c>$ are sampled according the to following generative model -- some context occurs in the world with probability $P(C=c)$. Within this context, an object $m$ occurs with probability $p(m|c)$. The speaker attempts to refer to that object by sampling from her conditional distribution $S(u|m, c; \ell)$ (i.e. $e \sim p(c)p(m|c)S(u|m, c; \ell)$). From the ingredients described above, it is possible to derive the following objective between the speaker and listener distributions (see \href{https://github.com/benpeloquin7/zipf_principles/blob/master/paper/supplementary_materials.pdf}{SI 2.1} for a complete derivation).

@@

From an information-theoretic perspective this objective is intuitive: $H_{cross}$ denotes the Cross-Entropy (CE), a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, we have an expectation over this term -- the expected difference between the distributions assumed by the speaker $P_{speaker}$ and listener $P_{listener}$ given a set of contexts $C$\footnote{Note that in the single context case $|C|=1$ this objective is simply the speaker-listener Cross-Entropy.}. In other words, an "efficient" language $\ell$ minimizes the distance between what speakers and listeners think.

@@

The task of understanding language is marked by a frequent need to handle various forms of ambiguity: lexical, syntactic, among others [@WasowPerforsBeaver2005a]. The ubiquity of this property, however, has been argued to provide evidence that languages have not been optimized for communication [@Chomsky2002a].

@@

In the following experiments we explore two aspects of Piantadosi’s et al.'s claim. In Simulation 1, we examine the efficient language *structure* aspect of their claim, exploring when the optimal linguistic system $\ell^*$ is most likely to contain ambiguous expressions. In Simulation 2, we explore an efficient language *use* aspect of the claim -- at what point in a conversation is it useful for a speaker to use ambiguous expressions?\par

```{r plot-optimal-langs, fig.env = "figure*", fig.pos = "h", fig.width=6, fig.height=2.8, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Panel (A) Vertical axis shows the proportion of optimal languages containing ambiguity. Horizontal axis shows the number of context-sizes (1-4). Optimal language under CE objective (red). Speaker-optimal (blue). Listener-optimal (green). Error bars represent 95 percent confidence intervals. Panel (B), example CE-optimal language (ambiguous) under in a four-condition context. Panel (C), example CE-optimal language (unambiguous) in a single-condition context."}
img <- png::readPNG("figs/fig1.png")
grid::grid.raster(img)
```

@@

We show that ambiguity is an efficient property under our CE objective in the reference game setting. We proceed by generating languages with different amounts of contextual support (varying the size of $|C|$). We search the space of languages, examining whether ones which minimize our objective contain ambiguity. If context leads to more efficient communication, then optimal languages should be more likely to be ambiguous as the amount of context increases.\par

@@

For comparison, we also examine properties of optimal languages under two additional objectives. Zipf [-@Zipf1949a] proposed that the speaker-optimal language $\ell_{speaker}^*$ would minimize speaker effort and the listener-optimal language $\ell_{listener}^*$ would minimize listener effort. We define these objectives using the first and second half of equation 1 (see \href{https://github.com/benpeloquin7/zipf_principles/blob/master/paper/supplementary_materials.pdf}{SI 2.2.}).\par

@@

In Simulation 1 we explored the degree to which ambiguity is an efficient property of languages when communication is contextualized. Figure 2, panel (A) plots the proportion of optimal languages under each objective as a function of number of contexts. The red line shows that as the number of contexts increases, so does the probability that an optimal language $\ell^*_{cross}$ contains ambiguity (at least one utterance maps to two meanings) under our CE objective. For comparison we also plot the proportion of speaker-optimal $\ell^*_{speaker}$ (blue line) and listener-optimal $\ell^*_{listener}$ (green line) languages that contain ambiguity. In line with Zipf's predictions, if languages are designed only to minimize speaker effort then optimal languages always contain ambiguity. If languages are designed to minimize listener effort then ambiguity is always avoided.\par

@@

We consider a single language $\ell$, which contains both ambiguous and unambiguous utterances. We assume ambiguous utterances are lower cost. Crucially, we do not assume that the listener knows the particular topic ($c_{current}$) of the conversation \textit{a priori}. Rather, that the listener has knowledge of the set of possible topics $C = \{c_1, \dots, c_k\}$, but does not know which one is currently being used by the speaker. Formally, this means the listener does not have access to the correct conditional distribution over meanings $P(M|C=c_{current})$ at the start of the discourse.\par

Over the course of a discourse $D$, the listener tries to infer both the current topic, $c_{current}$, as well as the particular meaning $m$ of a given utterance $u$. That is, we consider agents who can track the history of previous utterances $D$. Importantly, an agent can attempt to infer the current topic of conversation $c_{current}$ using the discourse history $D$.\par

@@

## Speaker agents
We vary the degree to which agents can use context for disambiguation. We consider three types of speaker models. Our \textit{Full pragmatics} agent, models a speaker who reasons about her listener and also has complete recall of the set of utterances in the discourse $D$. This speaker believes that the listener may not know the current topic $c_{current}$ at the start of the discourse, but can infer it over the discourse. We compare two baseline models. The first, a \textit{Partial pragmatics} baseline describes a speaker who reasons about a listener, but assumes they have no access to the discourse history. The second, a \textit{No pragmatics} baseline speaker does not consider a listener at all, but produces utterances according to the underlying langauge semantics ($\ell$) and topic probabilities ($p(M|C=c_{current}$) (see \href{https://github.com/benpeloquin7/zipf_principles/blob/master/paper/supplementary_materials.pdf}{SI 3}).\par

@@

Figure 2 (A) shows the empirical probability that a speaker uses an ambiguous utterance as a function of discourse position. The \textit{No pragmatic} baseline uses ambiguous utterances frequently and at a constant rate over the discourse. The \textit{Partial pragmatic} baseline avoids ambiguous utterances entirely. But, the \textit{Full pragmatic} model avoids ambiguous material only at the start of the discourse, employing it increasingly as the discourse proceeds. Panel (C) tracks our CE objective for each model over the discourse. Note that the objective decreases for all three models, primarily driven by the listener updating his belief about the actual topic ($P(C=c_{current}|D)$). However, the objective declines more quickly under the \textit{Full} and \textit{Partial} pragmatic speakers as listener agents are better able to infer the correct context. Additionally, the difference in CE between the \textit{Full}- and \textit{Partial pragmatic} models at the end of the discourse is driven by the reduction in speaker costs (see panel B). While speaker effort remains constant in both \textit{No pragmatic} and \textit{Partial pragmatic} baselines, effort declines in the \textit{Full pragmatic} model as she increasingly relies on ambiguous material later in the discourse.\par

@@

How do the competing pressures of speakers and listeners give rise to the distributional regularities found in natural language? Zipf [-@Zipf1949a] proposed that the asymmetry between speaker and listener costs gives rise to a range of properties at the level of the lexicon. We explored the interactions of rational pragmatic agents as a framework for understanding efficient language structure and use. We focused on an argument on the communicative function of ambiguity [@Piantadosi2011a], deriving a novel speaker-listener Cross-Entropy objective for measuring the efficiency of linguistic systems from first principles of efficient language use. In Simulation 1 we showed that optimal languages are more likely to contain ambiguous material when context is informative. In Simulation 2 we showed how rational pragmatic agents use ambiguous material efficiently in conversation.\par

A limitation of the current work is an analysis of exactly how the CE objective compares to existing measures. For example, previous work has described competing speaker-listener pressures in terms of a trade-off of simplicity and informativeness [@KempRegier2012a] or expressivity and compressibility [@SmithTamarizKirby2013a] to explain linguistic regularities. Future work should assess the degree to which we can derive the same properties as previous studies using our current framework. In line with this idea, a primary focus of the current work is to provide a case-study on the productivity of assessing functionalist theories in repeated reference games between rational pragmatic agents. We hope this framework can serve as a domain general tool to assess the range of functionalist theories examining efficient language-structure and use.\par

\vspace{1em} \fbox{\parbox[b][][c]{7.3cm}{\centering {Code and SI are available at this \href{https://github.com/benpeloquin7/zipf_principles/}{link}.}}} \vspace{1em}
\noindent
